Towards a standardized evaluation of imputation methodology

### 2022-07-27 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
After feedack Gerko on revisions

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently splits up the process of solving the missing data problem and the complete-data problem. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion), imputation allows analysts to account for the sources of uncertainty that are related to the missingness problem. With the popular variant multiple imputation \citep{rubi76}, each missing value is replaced by a plausible data value several times, resulting in multiple completed data sets. The estimated parameters of interest for each of these completed sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. [TODO: add that if the goal is statistical inference, the uncertainty due to missingness should be quantified/captured/propagated by the missing data method. Refer to \citet{sper20} for the context of prediction, where multiple imputation is not always necessary.] [TODO: paraphrase ``the quality of a technique can only be evaluated with respect to the aims of the problem/s it's intended to solve in practice".]

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. Such studies are typically composed of an analysis pipeline in which data are generated, made incomplete, and imputed repeatedly under different simulation conditions, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, there is no consensus on how to design, execute and report simulation studies aimed to evaluate imputation methodology. Without a `gold standard' for the evaluation of imputation routines, the validity and comparability of simulation setups may differ tremendously from one developer to another. This brings forth a chain of potential problems in the objective assessment of imputation method performance within and across studies, which may lead to sub-optimal use of imputation in practice.

@@

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]

Standardizing the evaluation of imputation methodology requires simulators to consider a couple of key aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology. [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

Within the scope of comparative simulation studies for the evaluation of imputation methodology for statistical inference, we observe several problems. To demonstrate the broad impact of these problems, we subdivide the problems into the following four distinct categories: problems with simulation design, problems with data generation, problems with missingness generation, and problems with performance evaluation. We further detail the impact each of these problems may have on the validity of evaluations.

@@

Extrapolating simulation results from a specific missing data generating model to more intricate (empirical) missingness could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may inadvertedly induce missingness according to the special case descibed by \citet[][p. \S 2.7]{buur18}. Under this specific missingness model, list-wise deletion may outperform any imputation method. Such a conclusion would, unfortunately, only translate to data that adhere to the same special case, while biasing inferences in other cases. % Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasinmissingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. 
Another example of potential side-effects of missingness generation may occur when multivariate missingness is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep[e.g. skewness;][]{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generation parameters.

@@

Missingness should be induced according to several sets of missing data conditions. We encourage simulators to consider different missingness patterns and mechanisms. Missingness mechanisms were first defined in Rubin's seminal work \citep{rubi76}. For more recent considerations see e.g. \citet{seam13, meal15, dore18, more18, moha21, scho21}. Although not every missingness mechanism is realistically assumed in practice, they can all offer valuable insights as simulation condition. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

[TODO: add segway to this paragraph and/or intrgrate into text above!] Every imputation workflow should contain an evaluation of the obtained imputations. Even though inspecting each imputation may be labor-intensive due to the number of imputations generated in a simulation study, we highly recommend to consider the following aspects:

@@

(iv) Finally, the plausibility of the imputed values may be evaluated. Plausible imputations--imputations that could be real values if they had been observed--are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausible imputations may be a desired property. One would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values. [TODO: add nuance here because valid inference, not plausible imputed values should be the aim.]

@@

Ultimately, this manuscript aims to elicit critical thinking about incomplete data simulation and to establish a common ground for the evaluation of imputation routines. Such a common ground would be the basis of a standardized evaluation. This would allow for fairer and more efficient comparisons between imputation techniques. Ultimately, it would be desirable to evaluate every imputation routine against the same standardized set in order to quantify the statistical properties across imputation routines. If properly executed, this would allow for careful matching of imputation methodologies to new missing data problems.

>>>>

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently separates the missing data problem from the analysis. When prediction is the goal, a single imputation may be sufficient \citet{sper20}. For inferential analyses, multiple imputation \citep{rubi76} has proven to be a valuable technique for obtaining valid inferences on incomplete datasets. In the case of multiple imputation each missing datum is imputed more than once, leading to multiple completed data sets. The resulting multiple inferences can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

% In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion), imputation allows analysts to account for the sources of uncertainty that are related to the missingness problem. With the popular variant multiple imputation \citep{rubi76}, each missing value is replaced by a plausible data value several times, resulting in multiple completed data sets. The estimated parameters of interest for each of these completed sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. [TODO: add that if the goal is statistical inference, the uncertainty due to missingness should be quantified/captured/propagated by the missing data method. Refer to \citet{sper20} for the context of prediction, where multiple imputation is not always necessary.] [TODO: paraphrase ``the quality of a technique can only be evaluated with respect to the aims of the problem/s it's intended to solve in practice".]

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. Such studies are typically composed of an analysis pipeline in which data are generated, made incomplete, and imputed repeatedly under different simulation conditions. A set of evaluation criteria is then postulated to evaluate the performance of one or more missing data methods. However, there is no consensus on how to design, execute and report simulation studies aimed to evaluate imputation methodology. Without a `gold standard' for the evaluation of imputation routines, the validity and comparability of simulation setups may differ tremendously from one developer to another. This brings forth a chain of potential problems in the objective assessment of imputation method performance within and across studies, which may lead to sub-optimal use of imputation in practice.

@@

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

Standardizing the evaluation of imputation methodology requires simulators to consider common aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology. [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

Within the scope of comparative simulation studies aimed at statistical inference, we observe several problems. To demonstrate the broad impact of these problems, we recognize the following four distinct categories: problems with simulation design, problems with data generation, problems with missingness generation, and problems with performance evaluation. We further detail the impact each of these problems may have on the validity of evaluations.

@@

Extrapolating simulation results from a specific missing data generating model to more intricate (empirical) missingness could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may inadvertedly induce missingness according to the special case descibed by \citet[][p. \S 2.7]{buur18}. Under this specific missingness model, list-wise deletion may outperform any imputation method. Such a conclusion would, unfortunately, only translate to data that adhere to the same special case, while biasing inferences in other cases. % Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasinmissingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. 
Another example of potential side-effects of missingness generation may occur when multivariate missingness is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generation parameters.

@@

Missingness should be induced according to several sets of missing data conditions. We encourage simulators to consider different missingness patterns and mechanisms. Missingness mechanisms were first defined in Rubin's seminal work \citep{rubi76}. For more recent considerations see e.g. \citet{seam13, meal15, dore18, more18, scho18, litt20, moha21, scho21}. Although not every missingness mechanism is realistically assumed in practice, they can all offer valuable insights as simulation condition. A disconnect between induced missingness, real missingness, and assumed missingness may result in simulation studies that are not as informative as they could be.

@@

[TODO: integrate this into text above!] Every imputation workflow should contain an evaluation of the obtained imputations. Even though inspecting each imputation may be labor-intensive due to the number of imputations generated in a simulation study, we highly recommend to consider the following aspects:

@@

(iv) Finally, the plausibility of the imputed values may be evaluated. Plausible imputations--imputations that could be real values if they had been observed--are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausible imputations may be a desired property. One would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values. [TODO: add nuance here because valid inference, not plausible imputed values, should be the aim (although plausibility is nice).]

@@

We aim to elicit critical thinking about incomplete data simulation and to establish a common ground for the evaluation of imputation routines. Such a common ground would be the basis of a standardized evaluation. This would allow for fairer and more efficient comparisons between imputation techniques. Ultimately, it would be desirable to evaluate every imputation routine against the same standardized set in order to quantify the statistical properties across imputation routines. If properly executed, such evaluations would allow for careful matching of imputation methodologies to new missing data problems.
